Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-04-08 mpaper/paper.tex
An example of notation and lots of editing

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}
\usepackage{thmtools, thm-restate}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage{layouts}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{csvsimple}
\usepackage{pgfplots}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{problem}[theorem]{Open Problem}

@@

\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\DKL}{D_{\mathrm{KL}}\infdivx}
\newcommand{\V}{V_{\mathbf{r}}}
\newcommand{\dx}{\,d\mathbf{r}\,d\mathbf{u}}
\newcommand{\vbound}{\frac{\rinf + \log|\mathcal{A}|}{1 - \gamma}}
\newcommand{\rinf}{\lVert \mathbf{r} \rVert_\infty}

@@

This learning problem
\cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} as well as
many others have benefited from an approach called \emph{inverse reinforcement
  learning} (IRL) (also known as inverse optimal control). IRL proposes a way to
learn behaviour from \emph{demonstrations} that typically come from human
actions. More formally, the IRL problem asks us to find a reward function for a
Markov decision process (MDP), where demonstrations are encoded as sets of
state-action pairs.

@@

% TODO: need to rewrite the first sentence
However, most IRL models in the literature make a convenient yet unjustified
assumption that the reward function can be expressed as a linear combination of
features
\cite{DBLP:conf/icml/PieterN04,DBLP:conf/icml/NgR00,ziebart2008maximum}.
This assumption makes the models unable to represent many reward structures. Out
of the non-linear models proposed to date, none can answer the questions posed
in the first paragraph. Quite often, the models assume that rewards have no
variance \cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}. In this paper,
we show how that assumption can be lifted by switching from maximum likelihood
estimation to \emph{variational inference} (VI), i.e., we approximate the
posterior distribution of the model by optimising the parameters of a simpler
distribution to make it similar to the posterior. This approach can prove useful
in four major ways:
\begin{enumerate}
\item By working with full distributions instead of point estimates, we can
  expect more precise reward predictions.
\item Variance estimates can be used to guide what data should be collected
  next, i.e., if the rewards of some states have abnormally high variance, we
  might want to expose the model to more data visiting those and surrounding
  states.
\item Variances estimates can also be used to judge whether we can trust the
  predictions of the model or, perhaps, the model could benefit from some
  adjustments or more data.
\item By adopting a more Bayesian approach, we automatically incorporate Occam's
  razor into the model that guards against overfitting
  \cite{DBLP:conf/uai/JinDAS17}.
\end{enumerate}

Our main contribution is a lengthy proof in Section \ref{sec:proof} that shows
how VI can be applied to the maximum-entropy IRL model with Gaussian processes
(GPs) proposed by Levine et al. \cite{DBLP:conf/nips/LevinePK11}. We describe
how we adapt the model and set up the VI problem in Section \ref{sec:model}.
Finally, in Section \ref{sec:evaluation} we examine the convergence properties
of our model and its ability to deduce optimal policies in practice.

@@

In this section, we introduce definitions and mathematical details relevant to
the problem. We begin by formally defining what the problem is.

\begin{definition}[MDP]
  A \emph{Markov decision process} is a set $\mathcal{M} = \{ \mathcal{S},
  \mathcal{A}, \mathcal{T}, \gamma, \mathbf{r} \}$, where $\mathcal{S}$ and
  $\mathcal{A}$ are sets of states and actions, respectively; $\mathcal{T} :
  \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is a function
  defined so that $\mathcal{T}(s, a, s')$ is the probability of moving to state
  $s'$ after taking action $a$ in state $s$; $\gamma \in [0, 1)$ is the discount
  factor; and $\mathbf{r} \in \mathbb{R}^{|\mathcal{S}|}$ is the reward
  vector\footnote{Depending on the situation, we will sometimes represent
    rewards as a function $r : \mathcal{S} \to \mathbb{R}$.}.
\end{definition}

\begin{definition}[IRL]
  Given an MDP without rewards $\mathcal{M} \setminus \{ \mathbf{r} \}$, an
  $|\mathcal{S}| \times d$ feature matrix $\mathbf{X}$ (where $d$ is the number
  of features), and a set of expert demonstrations $\mathcal{D} = \{\zeta_i
  \}_{i=1}^N$, where each demonstration $\zeta_i = \{ (s_{i,t}, a_{i,t})
  \}_{t=1}^T$ is a multiset of state-action pairs representing optimal actions
  executed by an expert, find the reward function that maximises the probability
  of observing the demonstrations, i.e.,
  \[
    \argmax_{\mathbf{r}} p(\mathcal{D} \mid \mathbf{r}).
  \]
\end{definition}

The optimal (deterministic) \emph{policy} $\pi : \mathcal{S} \to \mathcal{A}$
(i.e., a choice of actions for each state that maximises reward over time) is
usually constructed by defining a \emph{value (utility) function} $\V :
\mathcal{S} \to \mathbb{R}$ that measures how good a state is based on the
reward $\mathbf{r}$ as well as the structure of the MDP. One can then find $\V$
by applying the Bellman backup operator until convergence to every $s \in
\mathcal{S}$ (the technique is known as \emph{value iteration})
\cite{DBLP:books/daglib/0023820}:
\[
  V_{\mathbf{r}}(s) \coloneqq r(s) + \gamma \max_{a \in \mathcal{A}} \sum_{s' \in
    \mathcal{S}} \mathcal{T}(s, a, s')V_{\mathbf{r}}(s').
\]

@@

% TODO: rewrite the 'we write' sentence and the sentence before that
As we want learned rewards to generalise to previously unseen states and
for states to have a notion of similarity, the IRL definition also includes $d$
features associated with each state. In this paper, we will focus on modelling
rewards as a function from feature space to $\mathbb{R}$ using a \emph{Gaussian
  process}. A GP is defined as a collection of random variables, any finite
number of which has a joint Gaussian distribution
\cite{DBLP:books/lib/RasmussenW06}. We write $r \sim \mathcal{GP}(0, k)$ to say
that $r$ is a GP with mean $0$ and covariance function $k$. \emph{Covariance
  functions} (also known as \emph{kernels}) take two state feature vectors as
input and quantify how similar the two states are, in a sense that we would
expect high covariance scores to be associated with similar rewards.

@@

Our goal is then to use VI in order to approximate $\posterior$.
Let $\approximation$ denote our approximation. VI aims to optimise this
approximation by minimising the \emph{Kullback-Leibler} (KL) divergence between
the original probability distribution and our approximation. KL divergence
(asymmetrically) measures the difference between two probability distributions
and can be defined as \cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}_{\approximation}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}_{\approximation}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}_{\approximation}[\log p(\mathcal{D})].
\end{align*}
The last term is both hard to compute and constant with respect to
$\approximation$ \cite{blei2017variational}, so we can remove it from our
optimisation objective. The negation of what remains is known as the
\emph{evidence lower bound} (ELBO) and is defined as
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}_{\approximation}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iiint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}
Thus, instead of minimising KL divergence, we focus on maximising $\mathcal{L}$
by optimising the values of parameters to be defined in Section \ref{sec:model}.
Note that hereafter we drop the subscript notation, as all expected values will
be with respect to $\approximation$.

@@

We introduce a few definitions and results from linear algebra, numerical
analysis, and measure theory that will be used later in the paper. Namely, we
will use several different vector and matrix norms, consider how an inverse of a
matrix changes with a small perturbation, and introduce Lebesgue's dominated
convergence theorem which plays a major role in our theoretical results in
Section \ref{sec:proof}.

@@

\begin{theorem}[Dominated Convergence Theorem
  \cite{royden2010real}] \label{thm:lebesgue}
  Let $(X, \mathcal{M}, \mu)$ be a measure space and $\{ f_n \}$ a sequence of
  measurable functions on $X$ for which $\{ f_n \} \to f$ pointwise a.e. on $X$
  and the function $f$ is measurable. Assume there is a non-negative function
  $g$ that is integrable over $X$ and dominates the sequence $\{ f_n \}$ on $X$
  in the sense that
  \[
    |f_n| \le g \text{ a.e. on $X$ for all $n$.}
  \]
  Then $f$ is integrable over $X$ and
  \[
    \lim_{n \to \infty} \int_X f_n\,d\mu = \int_X f\,d\mu.
  \]
\end{theorem}

@@

In order to have a fully functional model, we still need to make several key
design decisions. In this section, we describe the covariance function used to
define the GP, consider what parameters should be used to optimise $\mathcal{L}$
and how they should be initialised, and in Section \ref{sec:elbo}, we derive our
final expression for $\mathcal{L}$.

@@

Ideally, we would like to model $\bm\lambda$ with an approximating distribution.
However, due to how the prior probability density function (PDF) for
$\mathbf{u}$ involves $\Kuu^{-1}$, and $\mathcal{L}$ has an expected value that
cannot be eliminated (see Section \ref{sec:elbo}), we are unable to show that
such an $\mathcal{L}$ is well-defined. More generally, we pose the following
problem, which, to the best of our knowledge, is currently open:
\begin{problem}
  Let $\mathbf{A}$ be a $n \times n$ matrix of coefficients, $X$ be a random
  variable, and $\mathbf{M}$ be an $n \times n$ matrix such that $M_{i,j} = f(X,
  A_{i,j})$, where $f$ is an arbitrary function. Under what circumstances does
  $\mathbb{E}[\mathbf{M}^{-1}]$ exist?
\end{problem}
While there are some obvious examples where the expected value exists (e.g.,
$f(X, A_{i,j}) = A_{i,j}X$ for an invertible $\mathbf{A}$ and many distributions
of $X$), it would be particularly interesting to know whether the answer is
`always'. A proof of such a result would allow us to model $\bm\lambda$ instead
of treating it as a variational parameter, and would help to guard against
overfitting. For now, $\bm\lambda$ will have to be treated as a variational
parameter.

@@

@@

The resulting model is summarised in Figure \ref{fig:graphical_model}. We rely
on $p(\mathcal{D} \mid \mathbf{r})$ as the only link between data and our model.
Since the expression for $q(\mathbf{r} \mid \mathbf{u})$ has both $\mathbf{u}$
and covariance matrices in it, $\mathbf{r}$ depends on both $\mathbf{u}$ and the
parameters of the kernel, $\bm\lambda$. The two remaining dependencies stem from
the fact that the approximating distribution for $\mathbf{u}$ is
$\mathcal{N}(\bm\mu, \mathbf{BB}^\intercal)$.

@@

Firstly, let us return to \eqref{eq:elbo} and write
\[
  \mathcal{L} = \mathbb{E}[\log\pfull] - \mathbb{E}[\log\approximation].
\]
By substituting in \eqref{eq:full} and \eqref{eq:approximation}, we get
\begin{align*}
  \mathcal{L} &= \mathbb{E}[\log p(\mathbf{u}) + \log p(\mathbf{r} \mid \mathbf{u}) + \log p(\mathcal{D} \mid \mathbf{r})] \\
              &- \mathbb{E}[\log q(\mathbf{u}) + \log q(\mathbf{r} \mid \mathbf{u})].
\end{align*}
Since $q(\mathbf{r} \mid \mathbf{u}) = p(\mathbf{r} \mid \mathbf{u})$, they
cancel each other out. Also notice that
\begin{align*}
  \mathbb{E}[\log p(\mathbf{u}) - \log q(\mathbf{u})] &= -\DKL{q(\mathbf{u})}{p(\mathbf{u})} \\
                                                      &= -\frac{1}{2} (\tr (\Kuu^{-1}\bm\Sigma) + \bm\mu^\intercal\Kuu^{-1}\bm\mu - m \\
                                                      &+ \log |\Kuu| - \log |\bm\Sigma|),
\end{align*}
by the definition of KL divergence between two multivariate Gaussians
\cite{kl}. Hence,
\begin{align*}
  \mathcal{L} &= \mathbb{E}\left[ \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right).
\end{align*}
Using the expressions for $Q_{\mathbf{r}}$ we get
\begin{align*}
  \mathcal{L} &= \mathbb{E}\left[\sum_{i=1}^N \sum_{t=1}^T r(s_{i,t}) - \V(s_{i,t}) + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t}, s')\V(s') \right] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right).
\end{align*}
We can simplify $\sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t})$ by defining a new vector
$\mathbf{t} = (t_1, \dots, t_{|\mathcal{S}|})^\intercal$, where $t_i$ is the
number of times the state associated with reward $r_i$ has been visited
across all demonstrations. Then,
\begin{align*}
  \mathbb{E} \left[ \sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t}) \right] &= \mathbb{E}[\mathbf{t}^\intercal\mathbf{r}] = \mathbf{t}^\intercal\mathbb{E}[\mathbf{r}] \\
                                                               &= \mathbf{t}^\intercal\mathbb{E}\left[\Kru^\intercal\Kuu^{-1}\mathbf{u}\right] = \mathbf{t}^\intercal\Kru^\intercal\Kuu^{-1}\bm\mu.
\end{align*}
This allows us to simplify $\mathcal{L}$ to
\begin{align*}
  \mathcal{L} &= \mathbf{t}^\intercal\Kru^\intercal\Kuu^{-1}\bm\mu - \mathbb{E}[v] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right),
\end{align*}
where
\[
  v = \sum_{i=1}^N \sum_{t=1}^T \V(s_{i,t}) - \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s_{i,t}, a_{i,t}, s')\V(s').
\]

>>>>

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}
\usepackage{thmtools, thm-restate}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage{layouts}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{csvsimple}
\usepackage{pgfplots}
\usepackage{blkarray}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{problem}[theorem]{Open Problem}
\newtheorem{example}[theorem]{Example}

@@

\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\DKL}{D_{\mathrm{KL}}\infdivx}
\newcommand{\V}{V_{\mathbf{r}}}
\newcommand{\dx}{\,\mathrm{d}\mathbf{r}\,\mathrm{d}\mathbf{u}}
\newcommand{\vbound}{\frac{\rinf + \log|\mathcal{A}|}{1 - \gamma}}
\newcommand{\rinf}{\lVert \mathbf{r} \rVert_\infty}

@@

This learning problem
\cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} as well as
many others have benefited from an approach called \emph{inverse reinforcement
  learning} (IRL) (also known as inverse optimal control). IRL proposes a way to
learn behaviour from \emph{demonstrations} that typically come from human
actions. More formally, the IRL problem asks us to find a reward function for a
Markov decision process (MDP), where demonstrations are encoded as sets of
state-action pairs.
% TODO (must): along with inverse optimal control, mention imitation learning
% and apprenticeship learning

@@

% TODO: need to rewrite the first sentence
However, most IRL models in the literature make a convenient yet unjustified
assumption that the reward function can be expressed as a linear combination of
features
\cite{DBLP:conf/icml/PieterN04,DBLP:conf/icml/NgR00,ziebart2008maximum}.
This assumption severely restricts what reward structures can be modelled (i.e.,
the properties of the function that represents rewards across states). Out of
the non-linear models proposed to date, none can answer the questions posed in
the first paragraph. Quite often, the models assume that rewards have no
variance \cite{DBLP:conf/nips/LevinePK11,DBLP:conf/uai/JinDAS17}.

In this paper, we show how that assumption can be lifted by switching from
maximum likelihood estimation to \emph{variational inference} (VI), i.e., we
approximate the posterior distribution of the model by optimising the parameters
of a simpler distribution to make it similar to the posterior. This approach can
prove useful in four major ways:
\begin{enumerate}
\item By working with full distributions instead of point estimates, we can
  expect more precise reward predictions.
\item Variance estimates can be used to guide what data should be collected
  next, i.e., if the rewards of some states have abnormally high variance, we
  might want to expose the model to more data visiting those and surrounding
  states.
\item Variances estimates can also be used to judge whether we can trust the
  predictions of the model or, perhaps, the model could benefit from some
  adjustments or more data.
\item By adopting a more Bayesian approach, we automatically incorporate Occam's
  razor into the model that guards against overfitting
  \cite{DBLP:conf/uai/JinDAS17}.
\end{enumerate}

Our main contribution is a proof in Section~\ref{sec:proof} that shows
how VI can be applied to the maximum-entropy IRL model with Gaussian processes
(GPs) proposed by Levine et al. \cite{DBLP:conf/nips/LevinePK11}. We describe
how we adapt the model and set up the VI problem in Section~\ref{sec:model}.
Finally, in Section~\ref{sec:evaluation} we examine the convergence properties
of our model and its ability to deduce optimal policies in practice.
% TODO: expand on this more

@@

In this section, we introduce definitions and mathematical details relevant to
the problem. We begin by formally defining the problem.

\begin{definition}[MDP]
  A \emph{Markov decision process} is a set $\mathcal{M} = \{ \mathcal{S},
  \mathcal{A}, \mathcal{T}, \gamma, \mathbf{r} \}$, where $\mathcal{S}$ and
  $\mathcal{A}$ are sets of states and actions, respectively; $\mathcal{T} :
  \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is a function
  defined so that $\mathcal{T}(s, a, s')$ is the probability of moving to state
  $s'$ after taking action $a$ in state $s$; $\gamma \in [0, 1)$ is the discount
  factor; and $\mathbf{r} \in \mathbb{R}^{|\mathcal{S}|}$ is the reward
  vector.
\end{definition}

While it is more common to represent rewards as a function $r : \mathcal{S} \to
\mathbb{R}$, the vector notation is generally more appropriate in our situation.
However, we switch between the two notations as needed.

\begin{definition}[IRL]
  Given an MDP without rewards $\mathcal{M} \setminus \{ \mathbf{r} \}$, an
  $|\mathcal{S}| \times d$ feature matrix $\mathbf{X}$ (where $d$ is the number
  of features), and a set of demonstrations $\mathcal{D} = \{\zeta_i
  \}_{i=1}^N$, where each demonstration $\zeta_i = \{ (s_{i,t}, a_{i,t})
  \}_{t=1}^T$ is a multiset of state-action pairs that represents optimal (or
  near-optimal) actions, find the reward function that maximises the probability
  of observing the demonstrations, i.e.,
  \[
    \argmax_{\mathbf{r}} p(\mathcal{D} \mid \mathbf{r}).
  \]
\end{definition}

\begin{example}
  Figure \ref{fig:example} shows a possible setup for IRL. In this case,
  \begin{align*}
    \mathcal{S} &= \{ s_1, s_2, s_3, s_4 \}, \\
    \mathcal{A} &= \{ l, r, u, d \}, \\
    \mathcal{T}(s, a, s') &= \begin{cases}
      1 & \text{if } (s, a, s') = (s_1, r, s_2) \\
      1 & \text{if } (s, a, s') = (s_1, d, s_4) \\
      1 & \text{if } (s, a, s') = (s_2, l, s_1) \\
      1 & \text{if } (s, a, s') = (s_2, d, s_3) \\
      1 & \text{if } (s, a, s') = (s_3, l, s_4) \\
      1 & \text{if } (s, a, s') = (s_3, u, s_2) \\
      1 & \text{if } (s, a, s') = (s_4, r, s_3) \\
      1 & \text{if } (s, a, s') = (s_4, u, s_1) \\
      0 & \text{otherwise,}
    \end{cases} \\
      \mathbf{X} &= \begin{blockarray}{ccc}
        & f_1 & f_2 \\
        \begin{block}{c[cc]}
          s_1 & 1 & 2 \\
          s_2 & 2 & 2 \\
          s_3 & 2 & 1 \\
          s_4 & 1 & 1 \\
        \end{block}
      \end{blockarray},
  \end{align*}
  and $\gamma \in [0, 1)$ can be assigned any value. The distribution of example
  state-action pairs into separate demonstrations is irrelevant to our model
  (and possible all other models), so we can keep each pair in its own set as
  follows:
  \[
    \mathcal{D} = \{ \{(s_1, d)\}, \{(s_2, l)\}, \{(s_4, u)\}, \{(s_4, u)\} \}.
  \]

  A solution to this problem is then an assignment of rewards to states, e.g.,
  \[
    \mathbf{r} = \begin{bmatrix}
      2 \\
      0 \\
      -1 \\
      1
    \end{bmatrix} \quad \text{or} \quad r(s) = \begin{cases}
      2 & \text{if } s = s_1 \\
      0 & \text{if } s = s_2 \\
      -1 & \text{if } s = s_3 \\
      1 & \text{if } s = s_4.
    \end{cases}
  \]
\end{example}

\begin{figure}
  \centering
  \begin{tikzpicture}[state/.style={circle,draw,fill=lightgray}]
    \node [state] (s1) {$s_1$};
    \node [state,right=2 of s1] (s2) {$s_2$};
    \node [state,below=2 of s2] (s3) {$s_3$};
    \node [state,left=2 of s3] (s4) {$s_4$};

    \draw[->,bend left,dashed] (s1) edge node[above] {$r$} (s2);
    \draw[->,bend left] (s2) edge node[below] {$l(1)$} (s1);
    \draw[->,bend left] (s1) edge node[right] {$d(1)$} (s4);
    \draw[->,bend left] (s4) edge node[left] {$u(2)$} (s1);
    \draw[->,bend left,dashed] (s4) edge node[above] {$r$} (s3);
    \draw[->,bend left,dashed] (s3) edge node[below] {$l$} (s4);
    \draw[->,bend left,dashed] (s2) edge node[right] {$d$} (s3);
    \draw[->,bend left,dashed] (s3) edge node[left] {$u$} (s2);

    \draw [<->] (-1.3, 0.8) node (yaxis) [above] {$f_2$}
    |- (3.5, -3.6) node (xaxis) [right] {$f_1$};
    \draw[shift={(0,-3.6)}] (0pt,2pt) -- (0pt,-2pt) node[below] {$1$};
    \draw[shift={(2.65,-3.6)}] (0pt,2pt) -- (0pt,-2pt) node[below] {$2$};
    \draw[shift={(-1.3,-2.6)}] (2pt,0pt) -- (-2pt,0pt) node[left] {$1$};
    \draw[shift={(-1.3,0)}] (2pt,0pt) -- (-2pt,0pt) node[left] {$2$};
  \end{tikzpicture}
  \caption{An example MDP with deterministic actions. States are represented by
    grey circles, actions by directed labelled edges between them. The numbers
    in parentheses next to some actions' names denote how many times that
    state-action pair appears in $\mathcal{D}$. If that number is zero, the edge
    is dashed. Finally, the axes assign two features to each state.}
  \label{fig:example}
\end{figure}

The set of demonstrations $\mathcal{D}$ often comes from observed human
behaviour and constitutes the most significant part of the data. The optimal
(deterministic) \emph{policy} $\pi : \mathcal{S} \to \mathcal{A}$ (i.e., a
choice of actions for each state that maximises reward over time) is usually
constructed by defining a \emph{value (utility) function} $\V : \mathcal{S} \to
\mathbb{R}$ that measures how good a state is based on the reward $\mathbf{r}$
as well as the structure of the MDP. One can then find $\V$ by applying the
Bellman backup operator until convergence to every $s \in \mathcal{S}$ (the
technique is known as \emph{value iteration}) \cite{DBLP:books/daglib/0023820}:
\[
  V_{\mathbf{r}}(s) \coloneqq r(s) + \gamma \max_{a \in \mathcal{A}} \sum_{s' \in
    \mathcal{S}} \mathcal{T}(s, a, s')V_{\mathbf{r}}(s').
\]

@@

A significant part of learning is about being able to generalise to previously
unseen input states (otherwise the problem becomes knowledge representation
instead of learning). For this reason, we represent each state by a
$d$-dimensional feature vector. This way, similar states are located in close
proximity to one another, and the ML model can predict rewards for new states
based on their similarity to training data.

In this paper, we model rewards as a function from feature space to
$\mathbb{R}$ using a \emph{Gaussian process}. A GP represents an infinite
collection of random variables, any finite number of which has a joint Gaussian
distribution \cite{DBLP:books/lib/RasmussenW06}. A GP is defined by its mean
function (which is always $\mathbf{x} \mapsto 0$ in our case) and covariance
function, which we denote by $k$. \emph{Covariance functions} (also known as
\emph{kernels}) take two state feature vectors as input and quantify how similar
the two states are, in a sense that we would expect high covariance scores to be
associated with similar rewards.

@@

Our goal is then to use VI in order to approximate
$p(\mathbf{u},~\mathbf{r}~\mid~\mathcal{D})$. Let $\approximation$ denote our
approximation. VI aims to optimise this approximation by minimising the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation. KL divergence (asymmetrically) measures the
difference between two probability distributions and is defined as
\cite{blei2017variational}
\begin{align*}
  \DKL{\approximation}{\posterior} &= \mathbb{E}_{\approximation}[\log\approximation - \log\posterior ] \\
                                   &= \mathbb{E}_{\approximation}[\log\approximation - \log\pfull] \\
                                   &+ \mathbb{E}_{\approximation}[\log p(\mathcal{D})].
\end{align*}
The last term is both hard to compute and constant with respect to
$\approximation$ \cite{blei2017variational}, so we can remove it from our
optimisation objective. The negation of what remains is known as the
\emph{evidence lower bound} (ELBO) and is defined as
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \mathbb{E}_{\approximation}\left[ \log \frac{\pfull}{\approximation} \right] \\
    &= \iint \log \frac{\pfull}{\approximation} \approximation\dx.
  \end{split}
\end{equation}
Thus, instead of minimising KL divergence, we focus on maximising $\mathcal{L}$
by optimising the values of parameters to be defined in Section~\ref{sec:model}.
Note that hereafter we drop the subscript notation, as all expected values will
be with respect to $\approximation$.

@@

We introduce a few definitions and results from linear algebra, numerical
analysis, and measure theory that will be used later in the paper. Namely, we
will use several different vector and matrix norms, consider how an inverse of a
matrix changes with a small perturbation, and introduce Lebesgue's dominated
convergence theorem which plays a major role in our theoretical results in
Section~\ref{sec:proof}.

@@

\begin{theorem}[Dominated Convergence Theorem
  \cite{royden2010real}] \label{thm:lebesgue}
  Let $(X, \mathcal{M}, \mu)$ be a measure space and $\{ f_n \}$ a sequence of
  measurable functions on $X$ for which $\{ f_n \} \to f$ pointwise a.e. on $X$
  and the function $f$ is measurable. Assume there is a non-negative function
  $g$ that is integrable over $X$ and dominates the sequence $\{ f_n \}$ on $X$
  in the sense that
  \[
    |f_n| \le g \text{ almost everywhere on $X$ for all $n$.}
  \]
  Then $f$ is integrable over $X$ and
  \[
    \lim_{n \to \infty} \int_X f_n\,\mathrm{d}\mu = \int_X f\,\mathrm{d}\mu.
  \]
\end{theorem}

@@

In order to have a fully functional model, we still need to make several key
design decisions. In this section, we describe the covariance function used to
define the GP, consider what parameters should be used to optimise $\mathcal{L}$
and how they should be initialised, and in Section~\ref{sec:elbo}, we derive our
final expression for $\mathcal{L}$.

@@

Ideally, we would like to model $\bm\lambda$ with an approximating distribution.
However, due to how the prior probability density function (PDF) for
$\mathbf{u}$ involves $\Kuu^{-1}$, and $\mathcal{L}$ has an expected value that
cannot be eliminated (see Section~\ref{sec:elbo}), we are unable to show that
such an $\mathcal{L}$ is well-defined. More generally, we pose the following
problem, which, to the best of our knowledge, is currently open:
\begin{problem}
  Let $\mathbf{A}$ be a $n \times n$ matrix of coefficients, $X$ be a random
  variable, and $\mathbf{M}$ be an $n \times n$ matrix such that $M_{i,j} = f(X,
  A_{i,j})$, where $f$ is an arbitrary function. Under what circumstances does
  $\mathbb{E}[\mathbf{M}^{-1}]$ exist?
\end{problem}
While there are some obvious examples where the expected value exists (e.g.,
$f(X, A_{i,j}) = A_{i,j}X$ for an invertible $\mathbf{A}$ and many distributions
of $X$), it would be particularly interesting to know whether the answer is
`always'. A proof of such a result would allow us to model $\bm\lambda$ instead
of treating it as a variational parameter, and would help to guard against
overfitting. For now, $\bm\lambda$ will have to be treated as a variational
parameter.

@@

The resulting model is summarised in Figure~\ref{fig:graphical_model}. We rely
on $p(\mathcal{D} \mid \mathbf{r})$ as the only link between data and our model.
Since the expression for $q(\mathbf{r} \mid \mathbf{u})$ has both $\mathbf{u}$
and covariance matrices in it, $\mathbf{r}$ depends on both $\mathbf{u}$ and the
parameters of the kernel, $\bm\lambda$. The two remaining dependencies stem from
the fact that the approximating distribution for $\mathbf{u}$ is
$\mathcal{N}(\bm\mu, \mathbf{BB}^\intercal)$.

@@

@@

Firstly, let us return to Equation~\eqref{eq:elbo} and write
\[
  \mathcal{L} = \mathbb{E}[\log\pfull] - \mathbb{E}[\log\approximation].
\]
By substituting in Equations~\eqref{eq:full} and \eqref{eq:approximation}, we get
\begin{align*}
  \mathcal{L} &= \mathbb{E}[\log p(\mathbf{u}) + \log p(\mathbf{r} \mid \mathbf{u}) + \log p(\mathcal{D} \mid \mathbf{r})] \\
              &- \mathbb{E}[\log q(\mathbf{u}) + \log q(\mathbf{r} \mid \mathbf{u})].
\end{align*}
Since $q(\mathbf{r} \mid \mathbf{u}) = p(\mathbf{r} \mid \mathbf{u})$, they
cancel each other out. Also notice that
\begin{align*}
  \mathbb{E}[\log p(\mathbf{u}) - \log q(\mathbf{u})] &= -\DKL{q(\mathbf{u})}{p(\mathbf{u})} \\
                                                      &= -\frac{1}{2} (\tr (\Kuu^{-1}\bm\Sigma) + \bm\mu^\intercal\Kuu^{-1}\bm\mu - m \\
                                                      &+ \log |\Kuu| - \log |\bm\Sigma|),
\end{align*}
by the definition of KL divergence between two multivariate Gaussians
\cite{kl}. Hence,
\begin{align*}
  \mathcal{L} &= \mathbb{E}\left[ \sum_{i=1}^N \sum_{t=1}^T Q_{\mathbf{r}}(s_{i,t}, a_{i,t}) - \V(s_{i,t}) \right] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right).
\end{align*}
Using the expressions for $Q_{\mathbf{r}}$ we get
\begin{align*}
  \mathcal{L} &= \mathbb{E}\left[\sum_{i=1}^N \sum_{t=1}^T r(s_{i,t}) - \V(s_{i,t}) + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t}, s')\V(s') \right] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right).
\end{align*}
We can simplify $\sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t})$ by defining a new vector
$\mathbf{t} = (t_1, \dots, t_{|\mathcal{S}|})^\intercal$, where $t_i$ is the
number of times the state associated with reward $r_i$ has been visited
across all demonstrations. Then,
\begin{align*}
  \mathbb{E} \left[ \sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t}) \right] &= \mathbb{E}[\mathbf{t}^\intercal\mathbf{r}] = \mathbf{t}^\intercal\mathbb{E}[\mathbf{r}] \\
                                                               &= \mathbf{t}^\intercal\mathbb{E}\left[\Kru^\intercal\Kuu^{-1}\mathbf{u}\right] = \mathbf{t}^\intercal\Kru^\intercal\Kuu^{-1}\bm\mu.
\end{align*}
This allows us to simplify $\mathcal{L}$ to
\begin{align*}
  \mathcal{L} &= \mathbf{t}^\intercal\Kru^\intercal\Kuu^{-1}\bm\mu - \mathbb{E}[v] \\
              &- \frac{1}{2} \left(\tr \left( \Kuu^{-1}\bm\Sigma \right) + \bm\mu^\intercal\Kuu^{-1}\bm\mu + \log |\Kuu| - \log |\bm\Sigma| \right),
\end{align*}
where
\[
  v = \sum_{i=1}^N \sum_{t=1}^T \V(s_{i,t}) - \gamma\sum_{s' \in \mathcal{S}}
  \mathcal{T}(s_{i,t}, a_{i,t}, s')\V(s').
\]
